
\input{preamble}

\title{Cohomology and base change, derived}
\author{Arpon Raksit}
\date{2017-02-09}

\numberwithin{block}{section}

\begin{document}
\maketitle

\newcommand{\QCoh}{\mathrm{QCoh}}

% ---------------------------------------------------------------------

\addtocounter{section}{-1}
\section{Introduction}

The goal of this note is to present the basic facts about and surrounding ``cohomology and base change'' theorems in algebraic geometry in light of the theory of derived algebraic geometry.

% ---------------------------------------------------------------------

\section{The derived results}

\begin{nothing}
  \label{derived-base-change}
  Suppose given a pullback diagram of derived schemes
  \[
    \begin{tikzcd}
      Y \ar[r, "q"] \ar[d, "g", swap] &
      X \ar[d, "f"] \\
      T \ar[r, "p"] &
      S.
    \end{tikzcd}
  \]
  There is a canonical map
  \[
    \beta \c p^*f_* \to g_*q^*
  \]
  of functors $\QCoh(X) \to \QCoh(T)$.

  \begin{subclaim}
    \label{derived-base-change-equivalence}
    Suppose $f$ (and therefore $g$) is quasicompact and quasiseparated. Then the map $\lambda$ defined above is an equivalence.
  \end{subclaim}
\end{nothing}

\begin{nothing}
  \label{derived-projection}
  Suppose given a morphism of derived schemes $f \c Y \to X$ and quasicoherent sheaves $\sF \in \QCoh(X)$ and $\sG \in \QCoh(Y)$. Then there is a canonical map
  \[
    \pi \c \sF \otimes f_*\sG \to f_*(f^*\sF \otimes \sG).
  \]

  \begin{subclaim}
    \label{derived-projection-formula}
    Suppose $f$ is quasicompact and quasiseparated. Then the map $\pi$ defined above is an equivalence.
  \end{subclaim}
\end{nothing}

\begin{claim}
  \label{derived-proper-pushforward}
  Let $f \c Y \to X$ be a morphism of derived schemes. Assume $f$ is proper and locally almost of finite presentation. Then $f_* \c \QCoh(Y) \to \QCoh(X)$ carries almost perfect objects to almost perfect objects.
\end{claim}

% ---------------------------------------------------------------------

% \bibliographystyle{amsalpha}
% \bibliography{refs}

\end{document}
